Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/lang.tex
Access syntax typo
Section 3 consistency with the new semantics

!= stencil backward(depth=2, dim=1) :: b
e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)

@@

!= stencil pointed(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
Not every dimension needs to be specified, \eg{},
specifications on lines $1$, $4$, and $10$ leave some dimensions unspecified.
The \term{forward}, \term{backward}, and \term{centered} regions may
all have an additional attribute \term{nonpointed} which marks absence
of the origin.  For example, the following is a
\term{nonpointed} \term{backward} stencil
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, nonpointed) :: a
b(i) = a(i-1) + 10*a(i-2)
\end{minted}
%
\paragraph{Combining regions}

The region operators \term{+} and \term{*} respectively combine
regions by union and intersection. The intersection of two regions
$r \term{*} s$ means that any indices in the specified code must be
consistent with both $r$ and $s$ simultaneously.
 Dually, for the union of two regions
 $r \term{+} s$ means that indices in the specified code must be
 inconsistent with one of $r$ or $s$, or both.
For example, the following \emph{nine-point stencil}
has a specification given by the product of two \texttt{centered}
regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%
%This pattern is common in image convolution applications.
The specification ranges over the
values that flow to the array subscript on the left-hand side,
and so ranges over the assignments to \term{x},
\term{y}, and \term{z}. Each index in the code is consistent
with both parts of the specification, \eg{}, \texttt{a(i-1, j+1)}
is within the centered region in dimension $1$ and the centered region
in dimension $2$.

@@

Region specifications can be assigned to region variables
(\emph{rvar}) via region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection
convolution~\cite{davis1975survey} can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common patterns, such as the five-point
pattern, as the regions can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in \Cref{fig:syntax}).
The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

Specifications can also be given to array computations which are not
stencils (\ie{}, do not have a left-hand side which is an array
subscript). For example, for a reduction:
%%
\begin{minted}{fortran}
do i = 1, n
  != access, pointed(dim=1) :: a
  r = max(a(i), r)
end do
\end{minted}
%%
The main difference between \fortran{stencil} and \fortran{access} is
that there is no left-hand side with which to check consistency or
relativise against.

>>>>

!= stencil backward(depth=2, dim=1) :: b
f(i) = b(i) + b(i-1) + b(i-2)

@@

!= stencil pointed(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
If a dimension contains only absolute index terms then it is left
unspecified.

The \term{forward}, \term{backward}, and \term{centered} regions may
all have an additional attribute \term{nonpointed} which marks absence
of the origin, corresponding to those array computations classified
as \textsf{straddle} in the previous section.  For example, the following is a
\term{nonpointed} \term{backward} stencil
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, nonpointed) :: a
b(i) = a(i-1) + 10*a(i-2)
\end{minted}
%
\paragraph{Combining regions}
The region operators \term{+} and \term{*} respectively combine
regions by union and intersection. The intersection of two regions
$r \term{*} s$ means that any indices in the specified code must be
consistent with both $r$ and $s$ simultaneously.
 Dually, for the union of two regions
 $r \term{+} s$ means that indices in the specified code must be
 consistent with one of $r$ or $s$, or both.
For example, the following \emph{nine-point stencil}
has a specification given by the product of two \texttt{centered}
regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%
%This pattern is common in image convolution applications.
The specification ranges over the
values that flow to the array subscript on the left-hand side,
and so ranges over the assignments to \term{x},
\term{y}, and \term{z}. Each index in the code is consistent
with both parts of the specification, \eg{}, \texttt{a(i-1, j+1)}
is within the centered region in dimension $1$ and the centered region
in dimension $2$.

@@

Region specifications can be assigned to region variables
(\emph{rvar}) via region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection
convolution~\cite{davis1975survey} can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common patterns, such as the five-point
pattern, as the regions can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in \Cref{fig:syntax}).
The \texttt{readOnce} modifier enforces that no subscript appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for a more informative and easily understood name
\texttt{readOnce}.

@@

Specifications can also be given to array computations which are not
stencils (\ie{}, do not have a left-hand side which is an array
subscript). For example, for a reduction:
%%
\begin{minted}{fortran}
do i = 1, n
  != access pointed(dim=1) :: a
  r = max(a(i), r)
end do
\end{minted}
%%
The main difference between \fortran{stencil} and \fortran{access} is
that there is no left-hand side with which to check consistency or
relativise against.

### 2017-04-18 oopsla17/study.tex
Section 3 consistency with the new semantics

\item Access patterns are mostly contiguous ($76.94\%$ of all array
  computations) and cross or straddle the origin ($75.75\%$), thus we
  can represent patterns by \emph{finite intervals} that meet at, or
  are next to, the origin.  These intervals will be given
  declaratively and abstractly, avoiding replicating indexing terms at
  the specification level.

\item We will support only consistent left and right-hand sides (where
  there are common induction variables on each side) ($66.01\%$)

>>>>

\item Access patterns are mostly contiguous ($76.94\%$ of all array
  computations) and cross or straddle the origin ($75.75\%$). We thus
  represent patterns by \emph{finite intervals} that meet at, or
  are next to, the origin.  These intervals will be given
  declaratively and abstractly, avoiding replicating indexing terms at
  the specification level.

\item We will support only consistent left and right-hand sides (where
  there are common induction variables on each side) ($66.01\%$ of all
  array computations).
